---
title: "Sars vs. COVID-19"
date: '2020-05-12'
author: "Melodie Irvin, mki77"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

Virology has always excited me, and with the coronavirus looming around everywhere we go, I found it compelling to compare the current data regarding the deaths and confirmed cases of COVID-19 with it's sister virus, SARS-CoV from 2003. Considering they are of the same genus and species of virus, I was expecting to find similar results regarding the number of deaths and rate of infection. As you will see, that is not at all the case. The Sars dataset was gathered from Kaggle while the COVID dataset was taken from ourworldindata.org. Both were combined in Excel prior to importing. The COVID dataset is up-to-date as of March 13, 2020. 


```{R}
options(repos="https://cran.rstudio.com" )
library(dplyr); library(formatR)
library(ggplot2)
library(readxl)
COVID_and_SARS <- read_excel("C:/Users/Melodie/Desktop/COVID_and_SARS.xlsx")
COVID_and_SARS<-COVID_and_SARS%>%slice(1:70)
COVID <-COVID_and_SARS%>%select(Date.Rank, Sum.of.COVID.Confirmed, Sum.of.COVID.Deaths)
SARS<-COVID_and_SARS%>%select(by=-c(Sum.of.COVID.Confirmed, Sum.of.COVID.Deaths))%>%
  select(Day=Date.Rank, Sum.of.Sars.Deaths, Sum.of.Sars.Confirmed)
```


```{R}
library(tidyverse)
FULL<- full_join(COVID, SARS, by=c("Date.Rank"="Day"))
FULL_2 <-FULL%>%mutate(
  Week.Number= case_when(
    between(`Date.Rank`,1,7)~"Week One",
    between(`Date.Rank`,8,14)~"Week Two",
    between(`Date.Rank`,15,21)~"Week Three",
    between(`Date.Rank`,22,28)~"Week Four",
    between(`Date.Rank`,29,35)~"Week Five",
    between(`Date.Rank`,36,42)~"Week Six",
    between(`Date.Rank`,43,49)~"Week Seven",
    between(`Date.Rank`,50,56)~"Week Eight",
    between(`Date.Rank`,57,63)~"Week Nine",
    between(`Date.Rank`,64,70)~"Week Ten",
    between(`Date.Rank`,71,77)~"Week Eleven",
    between(`Date.Rank`,78,84)~"Week Twelve",
    between(`Date.Rank`,85,91)~"Week Thirteen",
    between(`Date.Rank`,92,96)~"Week Fourteen"
  )
)%>%
  select(Date.Rank, Sum.of.COVID.Confirmed, Sum.of.Sars.Confirmed, Sum.of.COVID.Deaths,   Sum.of.Sars.Deaths, Week.Number)
FULL_New<-FULL_2%>%arrange(Date.Rank)%>%
  mutate(COVID.New.Cases=Sum.of.COVID.Confirmed-lag(Sum.of.COVID.Confirmed))%>%
  mutate(COVID.New.Deaths=Sum.of.COVID.Deaths-lag(Sum.of.COVID.Deaths))%>%
  mutate(Sars.New.Cases=Sum.of.Sars.Confirmed-lag(Sum.of.Sars.Confirmed))%>%
  mutate(Sars.New.Deaths= Sum.of.Sars.Deaths-lag(Sum.of.Sars.Deaths))%>%
  select(Date.Rank,Sum.of.COVID.Confirmed, Sum.of.Sars.Confirmed, Sum.of.COVID.Deaths, Sum.of.Sars.Deaths,
         COVID.New.Cases, Sars.New.Cases, COVID.New.Deaths, Sars.New.Deaths, Week.Number)
```

Given that this data is tidy, the pivot_longer and pivot_wider fuctions will be used in another section. The object of using those functions, however, is to organize data so that each observation has it's own row and each variable has it's own column. I chose to do a full_join of the data because I imported a single dataset and wanted to ensure I kept all components, therefore, no results were lost. A few additional columns were made for gathering summary statistics as well as to create a categorical variable. 

```{r}
FULL_New%>% filter(between(Date.Rank,25,50))
FULL_New%>% select(Sum.of.COVID.Confirmed,Sum.of.Sars.Confirmed, everything())%>%
  arrange(desc(Sum.of.COVID.Confirmed))
FULL_New%>%
  slice(1:53)%>%group_by(Week.Number)%>%
  summarize(mean_COVNew=mean(COVID.New.Cases,na.rm=T), sd_COVNew=sd(COVID.New.Cases, na.rm=T))
FULL_New%>%
  slice(1:53)%>%summarize(median(Sum.of.COVID.Confirmed))
FULL_New%>%
  summarize(median(Sum.of.Sars.Confirmed))
FULL_New%>%
  slice(1:53)%>%summarize(first(Sum.of.COVID.Deaths), last(Sum.of.COVID.Deaths), n_Week=n_distinct(Week.Number))
FULL_New%>%summarize(first(Sum.of.Sars.Deaths), last(Sum.of.Sars.Deaths), n_Week=n_distinct(Week.Number))
FULL_New%>%
  group_by(Week.Number)%>%
  summarize(mean_SARNew=mean(Sars.New.Cases,na.rm=T), sd_SARNew=sd(Sars.New.Cases, na.rm=T))
FULL_New%>%na.omit%>%summarize(cor(Sars.New.Cases, COVID.New.Cases))
FULL_New%>%
  slice(1:53)%>%summarize(mean_CNC=mean(COVID.New.Cases, na.rm=T), sd(COVID.New.Cases, na.rm=T))
FULL_New%>%na.omit%>%summarize(var(COVID.New.Cases, COVID.New.Deaths))

```

Given the amount of numeric varibles in the dataset, I only performed summary statistics on a few using the core dplyr functions. In order to find the difference in statistics on a set day, I filtered by date rank and found that the number of confirmed COVID cases on day 25 was more than 1500 percent of Sars cases on the same day in 2003. In addition, I selected the sums of COVID deaths and Sars deaths and arranged it in descending order of COVID deaths. Considering the COVID dataset was up-to-date as of March 13 (Day 53), I wanted to compare the values of both viruses. As of this time, the COVID death rate is around 3.73%, while the Sars virus from onset to conclusion resulted in around a 9.56% death rate. In other words, Sars was not as successful at spreading to as many hosts as COVID, but the people that contracted Sars had a higher chance of mortality than ones that have contracted COVID. The group_by function allowed me to group the categories based on week number and further summarize to find the mean and standard deviation of the variable regarding new cases of COVID per week. As shown, the highest average new cases for COVID occurred in the most recent week (week 8), while the highest for Sars came in week four. Considering some of the variable observations were accumulations, I took the median number for both regarding the sum of confirmed cases. COVID's median was 51,857 in 53 observations, while Sars median was 5,211 in 70 observations. This can be used as an indicator of the quantity of cases at half the time; days were used as the factor in this case. First and last sums of deaths were found to be four on day one and 774 on the last day for Sars virus. A correlation was done between new cases of Sars and new cases of COVID and was found to have a slighty negative correlation.

```{r}
library(tidyverse)
library(kableExtra)
COOR<-FULL_New%>%na.omit%>%select_if(is.numeric)
kable(COOR) %>%
  kable_styling(fixed_thead = T)
Tidycor<-cor(COOR)%>%as.data.frame%>%
  rownames_to_column%>%
  pivot_longer(-1,names_to="name",values_to="correlation")
head(Tidycor)
Tidycor%>%ggplot(aes(rowname,name,fill=correlation))+
  geom_tile()+
  scale_fill_gradient2(low="green",mid="purple",high="blue")+
  geom_text(aes(label=round(correlation,2)),color = "black", size = 4)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ 
  coord_fixed()
ggplot(FULL_New, aes(Sum.of.Sars.Deaths,Sum.of.Sars.Confirmed, color=Week.Number))+geom_point()+
  theme_light()+
  scale_x_continuous(breaks =seq(0,900, by = 100))+
  scale_y_continuous(lim=c(0,9000))+ 
  ggtitle("Cases and Deaths of Sars-CoV")
ggplot(FULL_New, aes(x=Sum.of.COVID.Confirmed, y=Sum.of.COVID.Deaths, fill=Week.Number)) +
  geom_violin(trim=F)+
  geom_boxplot(width=.1)+
  ggtitle("Cases and Deaths of COVID-19")
ggplot(FULL_New, aes(x = Week.Number))+
geom_bar(aes(y=COVID.New.Cases), stat="summary", fun.y="mean")+
  ggtitle("Average New COVID Cases per Week")
```

A coorelation heatmap calculates an integer to represent the relationship between variables. A number of "1" indicates sameness between the two variables and any coorelation close to that is considered to have a strong relationship. Some examples of this are seen between the sum of COVID confirmed cases and the date. However, some variables, such as the sum of COVID deaths and Sars new cases had a correlation of "0", meaning there is no relationship between the two. There were about equal distributions of strongly correlated variables as well as variables that had weak correlations. Only a few would be considered "moderate", as is seen between the sum of COVID confirmed and COVID new cases.

The cases and deaths of Sars virus were plotted with the color of dot indicating the week in which the deaths and cases occurred. The steepest slope which indicated the highest death rate per sum of cases happened around the onset of the virus and slowed toward its ending. There is a positive relationship between the two variables indicated by the positive slope of the line. As the sum of Sars deaths increased, the sum of confirmed cases increased.

Box and whisker plots show summary statistics between variables. The box itself indicates the IQR, representing the length between the lower and upper quartile, the middle line indicates the median value, and the whiskers show minimums and maximums. This third plot illustrated the sum of COVID confirmed cases and sum of COVID deaths. As suspected, there was a positive relationship between the two with a steady increase. As can be seen by the spacing in the third plot, the highest increase in deaths occurred between week seven and eight. 

The final plot used stat=summary to find the mean of new COVID cases by week number. Week one had the lowest average new cases, with week five coming in at second lowest. The highest average new case count came from week eight, with close to 6,000 new cases.

```{r}
library(cluster)
Clustered<-FULL_New%>%dplyr::select(-Week.Number)
Clustered1<-Clustered%>%na.omit%>%scale%>%as.data.frame()
sil_width<-vector() 
for(i in 2:10){
  kms <- kmeans(Clustered1,centers=i) 
  sil <- silhouette(kms$cluster,dist(Clustered1)) 
  sil_width[i]<-mean(sil[,3])
} 
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)
pam1<- Clustered1 %>% pam(k=2)
final<-Clustered1%>%mutate(cluster=as.factor(pam1$clustering))
confmat<-final%>%count(cluster)%>%arrange(desc(n))%>%
  pivot_wider(names_from="cluster",values_from="n",values_fill = list('n'=0))
confmat
ggplot(final, aes(x=Sum.of.COVID.Deaths,y=Sum.of.Sars.Deaths, color=cluster))+geom_point()

plot(pam1, which=2)
```

According to the ggplot, the highest average silhouette width came from five clusters (.463), but the second highest was very close and suggested two (.459). Given it is better to choose fewer clusters to achieve parsimony, I decided to have two. The quantity of clusters recommended was based on all numeric variables in the dataset, but only the sums of COVID deaths and Sars deaths were visualized. As seen on the ggplot, there were no real "clusters" or clear separation, but more evenly spaced out points throughout the graph. Given the silhouette width used to derive the clusters was not very high, the cluster solution achieved did not elicit much information. 
```{R}